Revisiting divisible, injective and flabby abelian groups from a constructive point of view

### 2022-09-29 slides-fischbachau2022.tex
Slides finished

  \normalsize
\end{frame}

>>>>

  \visible<3>{\emph{* higher-order proofs of first-order statements using the assumption \\
  that a given (perhaps uncountable) set is countable}}

  \pause
  \vspace*{1em}
  \mbox{\begin{minipage}{0.15\textwidth}
    \centering\small
    \includegraphics[height=5em]{zeta-function} \\
    $\mathbb{C}$
  \end{minipage}\quad
  \begin{minipage}{0.20\textwidth}
    \centering\small
    \includegraphics[height=5em]{3-adic-numbers} \\
    $\mathbb{Q}_p$
  \end{minipage}\quad
  \begin{minipage}{0.20\textwidth}
    \centering\small
    \includegraphics[height=5em]{bruyn-pope} \\
    $\mathbb{F}_1$
  \end{minipage}\quad
  \begin{minipage}{0.42\textwidth}
    \centering\small
    \includegraphics[height=5em]{hilbert-hotel} \\
    $\infty$
  \end{minipage}}
\end{frame}

\begin{frame}{A quantifier for finite approximations}
  Let~$X$ be a (perhaps uncountable) set. By a \hil{finite approximation} to a
  surjection~$\NN \twoheadrightarrow X$, we mean a \hil{finite list} of
  elements of~$X$. Notation:

  \begin{itemize}
    \item empty list: $[]$
    \item extension: $[x_1,\ldots,x_n] ::^r x_{n+1} = [x_1,\ldots,x_n,x_{n+1}]$
    \item refinement relation: $[x_1,\ldots,x_n,x_{n+1},x_{n+m}] \preceq [x_1,\ldots,x_n]$
    \item element access: $\sigma[i] = \text{element at position~$i$
    in~$\sigma$}$
  \end{itemize}
  \justifying

  For monotone predicates~$P$ of finite lists, we introduce
  a \hil{quantifier}~$\nabla$ such that~``$\nabla^{\tau \preceq \sigma}\_
  P(\tau)$'' expresses that \hil{no matter how~$\sigma$ evolves to a
  better approximation~$\tau$, eventually~$P(\tau)$ will hold}.
  \[
    \infer[{\text{\scriptsize$(\sigma \in X^*)$}}]{\nabla^{\tau \preceq \sigma}\_ P(\tau)}{P(\sigma)}
    \qquad
    \infer[{\text{\scriptsize$(\sigma \in X^*)$}}]{\nabla^{\tau \preceq
    \sigma}\_ P(\tau)}{\forall^{x \in X}\_\ \nabla^{\tau
    \preceq (\sigma ::^r x)}\_ P(\tau)}
  \]
  \[
    \infer[{\text{\scriptsize$(\sigma \in X^*, a \in X)$}}]{\nabla^{\tau
    \preceq \sigma}\_ P(\tau)}{\forall^{\tau \preceq \sigma}\_\
    a \in \tau \Rightarrow \nabla^{\upsilon \preceq \tau}\_ P(\upsilon)}
  \]
\end{frame}

\newcommand{\hnabla}{\hil{$\nabla$}}
\newcommand{\genalpha}{\mbox{$\hspace{0.12em}\shortmid\hspace{-0.62em}\alpha$}}

\begin{frame}{The generic surjection}
  \vspace*{-0.6em}
  \justifying
  \textbf{Def.} The \hil{$\nabla$-translation} $\varphi \mapsto
  \varphi^{\nabla}$ into formulas with a free variable $\sigma : X^*$ (denoting
  the \hil{current stage}) is defined by the following clauses.
  {\small
  \begin{align*}
    (\varphi_\text{atomic})^{\nabla} &\defeqv \hnabla^{\tau \preceq \sigma}\_
    \varphi_\text{atomic} \\
    \bot^{\nabla} &\defeqv \hnabla^{\tau\preceq\sigma}\_ \bot \\
    (\varphi \vee \psi)^{\nabla} &\defeqv \hnabla^{\tau\preceq\sigma}\_
      (\varphi^{\nabla}[\tau/\sigma] \vee \psi^{\nabla}[\tau/\sigma]) \\
    (\exists^{x\?X}\_ \varphi)^{\nabla} &\defeqv
    \hnabla^{\tau\preceq\sigma}\_(\exists^{x\?X}\_ \varphi^{\nabla}[\tau/\sigma]) \\
    (\varphi \Rightarrow \psi)^{\nabla} &\defeqv
      \forall^{\tau\preceq\sigma}\_ (\varphi^{\nabla}[\tau/\sigma] \Rightarrow
      \psi^{\nabla}[\tau/\sigma]) \\
    \top^{\nabla} &\defeqv \top \\
    (\varphi \wedge \psi)^{\nabla} &\defeqv (\varphi^{\nabla} \wedge \psi^{\nabla}) \\
    (\forall^{x\?X}\_ \varphi)^{\nabla} &\defeqv
      (\forall^{x\?X}\_
      \varphi^{\nabla}[\sigma/\tau]) \\
    (\genalpha(n){=}x)^\nabla &\defeqv
      (\hnabla^{\tau\preceq\sigma}\_ (\textsf{len}(\tau) > n \wedge \tau[n] = x))
  \end{align*}}

  \textbf{Ex.} $(\forall^{x\?X}\_ \exists^{n\?\NN}\_ \genalpha(n){=}x)^\nabla
  \equiv$ \\
  $\qquad\qquad
    (\forall^{x\?X}\_ \nabla^{\tau\preceq\sigma}\_
    \exists^{n\?\NN}\_
    \nabla^{\upsilon\preceq\tau}\_ (\textsf{len}(\upsilon) > n \wedge \upsilon[n] = x))$

  \textbf{Thm.} [Joyal--Tierney 1984] For \hil{first-order} formulas~$\varphi$,
  $\varphi^\nabla \Rightarrow \varphi$.
\end{frame}
